---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# ggmatplot

`ggmatplot` is a quick and easy way of plotting the columns of two matrices against each other. It is built upon [ggplot2](https://ggplot2.tidyverse.org/), and is aimed at providing a more direct and simple approach to creating plots using matrices.

`ggmatplot` can also be considered as an alternative to pivoting data into long format for visualizing using `ggplot2`.

## Installation

You can install the released version of `ggmatplot` from [GitHub](https://github.com/xuan-liang/ggmatplot):

```r
# install.packages("remotes")
remotes::install_github("xuan-liang/ggmatplot")
```

## Examples

```{r message=FALSE}
library(ggmatplot)
```

```{r}
USPersonalExpenditure

ggmatplot(
  rownames(USPersonalExpenditure),
  USPersonalExpenditure[, 1:5],
  plot_type = "point",
  xlab = "Category",
  ylab = "Expenditure (in Billions of Dollars)",
  legend_title = "Year",
  legend_label = c(1940, 1945, 1950, 1955, 1960)
) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
