---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%",
  message = FALSE
)
```
# ggmatplot

`ggmatplot` is a quick and easy way of plotting the columns of two matrices 
against each other. It is built upon [ggplot2](https://ggplot2.tidyverse.org/), 
and is aimed at providing a more direct and simple approach to creating plots 
using matrices.

`ggmatplot` can also be considered as an alternative to pivoting data into long 
format and visualizing using `ggplot2`.

## Installation

The latest version can be installed from [GitHub](https://github.com/xuan-liang/ggmatplot):

```r
# install.packages("remotes")
remotes::install_github("xuan-liang/ggmatplot")
```

## Examples

```{r}
library(ggmatplot)
library(ggplot2)
```

```{r point}
# matrix x
x <- rnorm(100,sd=2)

# matrix y
y = x*0.5+rnorm(100,sd=1)
fit.y = fitted(lm(y~x))
z <- cbind(y, fit.y)

ggmatplot(x, z)
```


```{r line}
# matrix x
x <- 1:10

# matrix y
y <- cbind(
  square = x^2,
  cube = x^3
)

ggmatplot(x, y,
          plot_type = "line",
          size = 1,
          color = c("blue", "purple"),
          ylim = c(0,750)) +
  theme_minimal()
```

```{r both}
USPersonalExpenditure

# matrix x
x <- rownames(USPersonalExpenditure)

# matrix y
y <- USPersonalExpenditure[, 1:5]

ggmatplot(x, y,
          plot_type = "both",
          xlab = "Category",
          ylab = "Expenditure (in Billions of Dollars)",
          legend_title = "Year",
          legend_label = c(1940, 1945, 1950, 1955, 1960)
          ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r density}
iris_sub <- subset(iris, Species == "setosa")

# matrix X
x <- (iris_sub[,1:2])
head(x,5)

ggmatplot(x, plot_type = "density") + 
  # scaling the density estimate to a maximum of 1
  aes(y = ..scaled..) +
  theme_bw()
```

```{r boxplot}
iris_sub <- subset(iris, Species == "setosa")

# matrix X
x <- (iris_sub[,1:4])
head(x,5)

ggmatplot(x, plot_type = "boxplot",
          alpha = 0, # removing fill values
          xlab = "", ylab = "")
```

```{r violin}
iris_sub <- subset(iris, Species == "setosa")

# matrix X
x <- (iris_sub[,1:2])
head(x,5)

ggmatplot(x, plot_type = "violin",
          color = c("#00AFBB", "#E7B800"),
          xlab = "", ylab = "")
```

```{r histogram}
iris_sub <- subset(iris, Species == "setosa")

# matrix X
x <- (iris_sub[,1:4])
head(x,5)

ggmatplot(x, plot_type = "histogram",
          xlab = "Group",
          color = "black",
          fill = c("#F8766D", "#7CAE00", "#00BFC4", "#C77CFF"))
```



