---
title: "Introduction to `ggmatplot`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggmatplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggmatplot)
library(ggplot2)
```

The `ggmatplot` package provides a quick and easy way of plotting the columns of two matrices against each other using [`ggplot2`](https://ggplot2.tidyverse.org/).

This vignette is aimed at providing a brief understanding of how the `ggmatplot()` function works and possible extensions to the plots created using `ggmatplot`. This will be discussed through the following examples.


## `point` plot

First we create two matrices as follows:

`x`: a vector/a single columned matrix

```{r point-plot-x}
# matrix x
x <- rnorm(100, sd = 2)
head(x)
```

`y`: a matrix with two columns - `z` and `fit.z`

```{r point-plot-y}
# matrix y
z <- x * 0.5 + rnorm(100, sd = 1)
fit.z <- fitted(lm(z ~ x))
y <- cbind(z, fit.z)
head(y)
```

`ggmatplot()` can be used to plot `x` against each column of `y` - `x` against `z`, and `x` against `fit.z`. This will be represented on the resulting plot as two groups, identified using different shapes and colors. 

```{r basic-point-plot}
ggmatplot(x, y)
```

The default aesthetics used to differentiate the two groups can be updated using the `ggmatplot()` function parameters. Since the two groups in this example are differentiated using their shapes and colors, the `shape` and `color` parameters can be used to change them. If we want points in both groups to have the same shape, we can simply set the `shape` parameter to a single value. However, if we want the points in the groups to be differentiated by color, we can pass a list of colors as the `color` parameter - but we should make sure the number of colors in the list matches up with the number of groups.

```{r point-plot-w-parameters}
ggmatplot(x, y,
          shape = "circle", # using single shape over both groups
          color = c("blue","purple") # assigning two colors to thw two groups
          )
```

Since `ggmatplot` is built upon `ggplot2` and creates a ggplot object, add on layers such as scales, themes and coordinate systems that can be layered over ggplot objects, can be used with plots created using `ggmatplot` too.

Each `plot_type` allowed by the `ggmatplot()` function is also built upon a `ggplot2` geom(geometric object), and will therefore support additional parameters accepted by the underlying geom, although they are not listed as `ggmatplot()` function parameters.

```{r point-plot-w-layers}
ggmatplot(x, y,
          shape = "circle", # using single shape over both groups
          color = c("blue","purple"), # assigning new colors to groups
          size = 2,
          alpha = 0.5
          ) +
  facet_wrap(~Group) +
  theme_bw()
```


## `line` plot

We have the following dataset of the monthly totals of international airline passengers(in thousands) from January 1949 to December 1960.

```{r line-data}
AirPassengers <- matrix(AirPassengers,
  ncol = 12, byrow = FALSE,
  dimnames = list(month.abb, as.character(1949:1960))
)

AirPassengers
```
If we want to plot the trend in the number of passengers over the years, we could break the data down into two matrices as follows:

`months`: a vector/single columned matrix containing the list of months  
`nPassengers`: a matrix of passenger numbers with each column representing a year

```{r line-data-split}
months <- rownames(AirPassengers)
nPassengers <- AirPassengers[, 1:12]
```

Then we can use `ggmatplot()` to plot the `months` matrix against each column of the `nPassengers` matrix - which can be more simply understood as grouping the plot using each column(`year`) of the `nPassengers` matrix.

```{r line-plot}
ggmatplot(
  x = months,
  y = nPassengers,
  plot_type = "line",
  size = 1,
  legend_label = c(1949:1960),
  xlab = "Month",
  ylab = "Total airline passengers(in thousands)",
  legend_title = "Year"
) +
  theme_minimal()
```

## `density` plot

Next we use a subset of the `iris` dataset to plot out the kernel density estimates of each variable. The data corresponds to 4 different measurements of the iris flowers.

```{r density-data}
iris_setosa <- subset(iris, Species == "setosa")

head(iris_setosa[,1:4])
```

The plot types `density`, `histogram`, `violin` and `boxplot` only accept a single matrix input. The plot will be grouped based on the columns of the single matrix input. Since we used a matrix with 4 columns, we can find 4 groups in the following density plot.

A transparent fill can be achieved by setting `alpha = 0`.

```{r density-basic}
ggmatplot(iris_setosa[,1:4],
          plot_type = "density",
          xlab = "",
          alpha = 0,
          size = 1)
```

If we want to scale the y axis to a maximum of 1, it can be achieved as follows:

```{r density-scaled}
ggmatplot(iris_setosa[,1:4],
          plot_type = "density",
          xlab = "") +
  aes(y = ..scaled..)
```

Parameters accepted by the underlying geom - `geom_density` can also be used within the `ggmatplot()` function when using `plot_type = "density"`.

```{r density-stacked}
ggmatplot(iris_setosa[,1:4],
          plot_type = "density",
          xlab = "",
          position = "stack")
```
